Isomorphic Interpreters from Logically Reversible Abstract Machines

### 2012-03-16 iso-int/rc-iso-int.tex
sizes
many small fixes

\item {{fold}} and {{unfold}} are basically drawn as triangular boxes
  that map the a recursive type to its unfolding.

\item {{trace+}} is drawn as a looped circuit, where the traced value
  is shown as flowing backward. The semantics and interpretation of
  {{trace}} are dealt with in detail in previous work (see Sec. 4 of
  \cite{infeffects}).

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Simple Bounded Number Iteration}

@@

\begin{multicols}{2}
%subcode{bnf} include main
% Numbers, n, m = 0 | n + 1
% Machine states = <n, n>

%subcode{bnf} include main
% Start state = <n, 0>
% Stop State = <0, n>
\end{multicols}
% Small-step operational semantics:

%subcode{opsem} include main
% <n+1, m> |--> <n, m+1>

@@

A closer look however reveals that the machine defines a \emph{partial}
isomorphism: not all valid final states can be mapped to valid start
states. Indeed consider the configuration {{<0,2,3>}} which is a valid final
state. Going backwards, the transitions start as follows 
{{<0,2,3> |--> <1,1,2> |--> <2,0,1>}} at which point, the machine gets stuck 
at a state that is not a valid start state.  There are several ways to deal with this
situation. The first is to use a partial isomorphism that diverges on the stuck state.
The second is to add another
possible kind of start state {{<n,0,n>}} to make the isomorphism total. 
Assuming the seond solution, our construction now proceeds as
expected. After expanding the input to expose $n$ and the output to expose
both $p$ and $m$, we reach the following configuration:

@@

\begin{center}
\scalebox{0.7}{
  \includegraphics{diagrams/adder3.pdf}
}
\end{center}

We have also connected the matching input and output wires. At this point, we
have exposed both start states and the stop state. 
% \begin{enumerate}
% \item We can hide the use the combinator {{omega : b <-> 0}}
%   expressible in {{langRevT}}. This combinator is basically an
%   infinite loop on all inputs and hence would allow us to entirely
%   discard one possible branch. In other words, we are explicitly
%   admitting that our machine is undefined on some inputs.
% \item We can admit that there are two possible start states to this
%   abstract machine {{<n, m, 0>}} and {{<n,0,m>}}. One of this is an
%   artifact of the fact that backward execution encounters stuck
%   states. Dually, if there were stuck states for forward execution,
%   they would be reflected in the fact that there were multiple stop
%   states -- only one of which denotes a desirable or correct execution
%   of the machine.
% \end{enumerate}
We can slide the two sides of the diagram and tie the knot using {{trace}} to
get the desired combinator:

\begin{center}
\scalebox{0.7}{
  \includegraphics{diagrams/adder4.pdf}
}
\end{center}

@@

%%%%%%%%%%%%%%%%%%%%%
\subsection{A {{langRevT}} Interpreter}

@@

\begin{small}

@@

\end{small}

The isomorphic interpreter thus derived is shown below, just before
the `flipping' step (which makes it harder to read). Each wire has
been annotated with the reduction rule it corresponds to making it
easier to compare with the abstract machine.

@@

\noindent
A couple of important points:

@@

\item The abstract machine definition is not very strongly typed in
  teh following sense: we know that every {{v:b1*b2}} must have the
  form {{(v1,v2)}} however is cases like {{rule 8}} if we allow any
  {{v}} to appear the machine will be in an undefined state. The usual
  way this is handled in any interpreter is to either test to ensure
  that the runtime value has teh required form or to user s stronger
  typing discipline such as GADTs.We abstract away from this choice in
  this presentation.
\end{enumerate}

@@

\bibliographystyle{splncs03} 
\bibliography{cites}
\end{document}

>>>>

\item Operations {{fold}} and {{unfold}} are specific to the recursive
  data type and are drawn as triangular boxes. Sec. \ref{sec:bounded}
  for instance shows the {{fold}}/{{unfold}} for recursive type of
  natural numbers, {{nat=@@x.(1+x)}}.

\item The {{trace+}} operation is drawn as a looped circuit, where the
  traced value is shown as flowing backward. The semantics and
  interpretation of {{trace}} are dealt with in detail in previous
  work (see Sec. 4 of \cite{infeffects}).

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Simple Bounded Number Iteration}
\label{sec:bounded}

@@

\vspace{-15pt}
\begin{multicols}{2}
%subcode{bnf} include main
% Numbers, n, m = 0 | n + 1
% Machine states = <n, n>

%subcode{bnf} include main
% Start state = <n, 0>
% Stop State = <0, n>
\end{multicols}
\vspace{-15pt}
%subcode{opsem} include main
% <n+1, m> |--> <n, m+1>

@@

A closer look however reveals that the machine defines a
\emph{partial} isomorphism: not all valid final states can be mapped
to valid start states. Indeed consider the configuration {{<0,2,3>}}
which is a valid final state. Going backwards, the transitions start
as follows {{<0,2,3> |--> <1,1,2> |--> <2,0,1>}} at which point, the
machine gets stuck at a state that is not a valid start state.

\paragraph*{Stuck States.}
In a more general sense, all abstract machines have to deal with stuch
states and there are several standard ways of delaing with them:
\begin{itemize}
\item \emph{Use better types.} One could augment {{langRevT}} with a
  richer type system that distinguishes non-zero numbers from those
  that can be zero, thus emilinating this situation entirely. In the
  intepreter in Sec. \ref{sec:langrevt-int} for instance one could use
  a GADT to eliminate stuck states.

\item \emph{Diverge.} Another standard approach in dealing with stuck
  states is to make the diverge or leave the output
  undefined/unobservable in some way. For instance {{add1}} defined in
  {{langRevT}} (see Sec. 4.2 in \cite{James:2012:IE:2103656.2103667})
  has teh property that {{add1^{dagger} 0}} diverges.

\item \emph{Stop the machine.} In the case of our machine this
  corresponds to another possible kind of start state {{<n,0,n>}} to
  make the isomorphism total.  Let us choose this path in the case of
  this machine.  After expanding the input to expose $n$ and the
  output to expose both $p$ and $m$, we wire the machine corresponding
  to the reduction rule.

\begin{center}
\scalebox{0.8}{
  \includegraphics{diagrams/adder3.pdf}
}
\end{center}

Note that there are indeed two start states in the interpreter. We can
slide the two sides of the diagram and tie the knot using {{trace}} to
get the desired interpreter:

\begin{center}
\scalebox{0.8}{
  \includegraphics{diagrams/adder4.pdf}
}
\end{center}

%% We have also connected the matching input and output wires. At this
%% point, we have exposed both start states and the stop state.

%% There are several ways to deal with this situation. The first is to
%% use a partial isomorphism that diverges on the stuck state.

@@

% \begin{enumerate}
% \item We can hide the use the combinator {{omega : b <-> 0}}
%   expressible in {{langRevT}}. This combinator is basically an
%   infinite loop on all inputs and hence would allow us to entirely
%   discard one possible branch. In other words, we are explicitly
%   admitting that our machine is undefined on some inputs.
% \item We can admit that there are two possible start states to this
%   abstract machine {{<n, m, 0>}} and {{<n,0,m>}}. One of this is an
%   artifact of the fact that backward execution encounters stuck
%   states. Dually, if there were stuck states for forward execution,
%   they would be reflected in the fact that there were multiple stop
%   states -- only one of which denotes a desirable or correct execution
%   of the machine.
% \end{enumerate}

@@

%%%%%%%%%%%%%%%%%%%%%
\subsection{A {{langRevT}} Interpreter}
\label{sec:langrevt-int}

@@

\begin{scriptsize}

@@

\end{scriptsize}

@@

The isomorphic interpreter thus derived is shown below, just before
the `flipping' step (which makes it harder to read). Each wire has
been annotated with the reduction rule it corresponds to making it
easier to compare with the abstract machine.  A couple of important
points:

@@

\item Stuck states in this machine correspond to values not matching
  the shape expected by their types. They can be handled as decribed
  in Sec. \ref{sec:adder} using GADTs (which eliminates them
  entirely), divergence, or adding extra halting states.

@@

\begin{scriptsize}
\bibliographystyle{splncs03} 
\bibliography{cites}  
\end{scriptsize}
\end{document}
